Data Descriptor for Nature Scientific Data

### 2024-11-27 manuscript/manuscript-nature/manuscript-nature.qmd
manuscript-and-fig-cities

# Main {.unnumbered}

<!--# CCA: Fr to rewrite section -->

@@

# Results

## Analysing the variability in the impact of COVID-19 on mobility across population groups

@@

## Affluent and densely populated areas saw steepest decline in daily mobility, but all population groups progress towards baseline levels

@@

Focusing first on the patterns by population density category, we generally observe a decrease in the number of inflows for all countries and most categories of analysis in the early pandemic days.
The extent of this decrease tends to be larger for high-density areas.
Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density.
In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage changes exceeding 100%.
However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage changes.
Nonetheless, these percentage changes illustrate the extent of the impacts experienced by these sparsely populated regions.
In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories.
Notably, mobility levels in areas characterised by higher population density still remain below the baseline.
Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline.
These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May).
Overall, the observed patterns by population density category are consistent with those reported in CITE for Argentina and Chile, where the data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.

@@

These patterns can be attributed to...
THIS GOES TO THE DISCUSSION

@@

## Initial drop in mobility drives intergroup variation in mobility levels, as recovery pace is uniform across population groups

Next we focus on modelling intergroup differences in mobility trends over time, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups.
@fig-recovery-density and @fig-recovery-rdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively.
Time is represented on the $x$-axis, spanning from mid-April 2020 to mid-April 2022, while the intensity of movement is depicted on the $y$-axis, measured as the percentage change in the number of flows relative to pre-pandemic baseline levels.
The background of each plot in Figures @fig-recovery-density and @fig-recovery-rdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns.
This data provides context for the observed trends, reflecting the mobility restrictions imposed by each country during the specified periods.

We highlight four key observations from a visual inspection of the evolution patterns by population density category, as shown in @fig-recovery-density.
First, there is a general upward trend in mobility across all population density categories for all countries, reflecting a recovery of mobility levels following the early days of COVID-19.
Second, more densely populated areas consistently record lower levels of mobility throughout the study period.
Third, there exist some variations in the patterns by country.
Chile experiences greater overall losses in mobility compared to Argentina, while certain areas in Argentina reach levels comparable to the pre-pandemic baseline.
In contrast, Colombia displays more varied trends across population density categories, with above-baseline levels of mobility in low-density areas, and significantly larger losses in high-density areas.
Fourth, the temporal patterns display fluctuations from the general trend.
These fluctuations are likely driven by seasonality effects and by the changing levels of stringency of COVID-19 interventions.

![](figures/Fig-recovery-density.png){#fig-recovery-density}

Visual inspection of @Fig-recovery-rdi reveals the following four patterns by relative deprivation index category.
First, we observe a general upward trend in mobility across all relative deprivation index categories for all countries.
Second, less deprived areas consistently record lower levels of mobility throughout the study period.
In contrast, more deprived areas exhibit higher levels of mobility.
Third, there is variability by country.
Colombia stands out for recording mobility levels in the most deprived areas that are well above the baseline.
These big relative changes for the most deprived category does not necessarily correspond to large population numbers since for example, in Colombia, more deprived regions are often low-density areas that typically contain small populations.
As a result, minor fluctuations in mobility can lead to substantial percentage changes.
Nevertheless, the positive percentage changes observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.
Fourth

![](figures/Fig-recovery-rdi.png){#fig-recovery-rdi}

Next, we quantitatively model intergroup differences in mobility trends over @fig-recovery-density and @fig-recovery-rdi.
To achieve this, we perform a time series decomposition into three components: trend, seasonal, and residual.
Our analysis focuses on the trend component, which we model using five specifications within a linear mixed-effects modeling framework, all yielding consistent results.
All models include time as a fixed effect and estimate an intercept and a slope associated with this variable.
The intercept is interpreted as the magnitude of the initial drop in mobility, while the slope represents the rate at which mobility levels recover over time.

Models 1 and 2 consider only fixed effects.
Models 3, 4, and 5 incorporate time as a fixed effect while also accounting for random effects to capture variation in the intercept, slope, or both, based on the category of the origin of the flows.
Further details on the model specifications are provided in Methodology Section X. Figures @fig-re-density and @fig-re-rdi present the random variations in intercept and slope estimated by Models 3, 4, and 5, which provide insights into the drivers of intergroup variation in the recovery towards baseline mobility levels.

![](figures/Fig-re-density.pdf){#fig-re-density}

![](figures/Fig-re-rdi.pdf){#fig-re-rdi}

INTERPRET

>>>>

# Main {#sec-introduction}

*`@Francisco to rewrite this section`*

@@

# Results {#sec-results}

@@

## Affluent and densely populated areas saw steepest decline in daily mobility, but all population groups progress towards baseline levels {#sec-resultsevooverview}

@@

Focusing first on the patterns by population density category, we generally observe a decrease in the number of inflows for all countries and most categories of analysis in the early pandemic days.
The extent of this decrease tends to be larger for high-density areas.
Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density.
In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage changes exceeding 100%.
However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage changes.
Nonetheless, these percentage changes illustrate the extent of the impacts experienced by these sparsely populated regions.
In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories.
Notably, mobility levels in areas characterised by higher population density still remain below the baseline.
Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline.
These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May).
Overall, the observed patterns by population density category are consistent with those reported in [@RoweCCA24] for Argentina and Chile, where data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.

@@

*`@Miguel, can you elaborate the following in the discussion? - The pattern that we observe in this section can be attributed to... THIS GOES TO THE DISCUSSION`*

@@

## Initial drop in mobility drives intergroup variation in mobility levels, as recovery pace is uniform across population groups {#sec-resultsintergrouprecovery}

Next we focus on modelling intergroup differences in mobility trends over time, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups.
@fig-recoverydensity and @fig-recoveryrdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively.
Time is represented on the $x$-axis, spanning from mid-April 2020 to mid-April 2022, while the intensity of movement is depicted on the $y$-axis, measured as the percentage change in the number of flows relative to pre-pandemic baseline levels.
The background of each plot in Figures @fig-recoverydensity and @fig-recoveryrdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns.
This data provides context for the observed trends, reflecting the mobility restrictions imposed by each country during the specified periods.

*`@Francisco and @Miguel to check that what I wrote below makes sense`*

We highlight four key observations from a visual inspection of the evolution patterns by population density category, as shown in @fig-recoverydensity. First, there is a general upward trend in mobility across all population density categories for all countries, reflecting a recovery of mobility levels following the early days of COVID-19.
Second, more densely populated areas consistently record lower levels of mobility throughout the study period.
Third, there exist some variations in the patterns by country.
Chile experiences greater overall losses in mobility compared to Argentina, while certain areas in Argentina reach levels comparable to the pre-pandemic baseline.
In contrast, Colombia displays more varied trends across population density categories, with above-baseline levels of mobility in low-density areas, and significantly larger losses in high-density areas.
Fourth, the temporal patterns display fluctuations from the general trend.
These fluctuations are likely driven by seasonality effects and by the changing levels of stringency of COVID-19 interventions.

![Figures](figures/Fig-recovery-density.png){#fig-recoverydensity}

Visual inspection of @fig-recoveryrdi reveals the following four patterns by relative deprivation index category.
First, we observe a general upward trend in mobility across all relative deprivation index categories for all countries.
Second, less deprived areas consistently record lower levels of mobility throughout the study period.
In contrast, more deprived areas exhibit higher levels of mobility.
Third, there is variability by country.
Colombia stands out for recording mobility levels in the most deprived areas that are well above the baseline.
These big relative changes for the most deprived category do not necessarily correspond to large population numbers since for example, more deprived regions in Colombia are often low-density areas that typically contain small populations.
As a result, minor fluctuations in mobility can lead to substantial percentage changes.
Nevertheless, the positive percentage change observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.
Fourth, like in @fig-recoverydensity, the temporal patterns display fluctuations from the general trend which are likely driven by seasonality effects and by the changing levels of COVID-19 stringency index.

![Figures RDI](figures/Fig-recovery-rdi.png){#fig-recoveryrdi}

Next, we quantitatively model intergroup differences in mobility trends over @fig-recoverydensity and @fig-recoveryrdi. To achieve this, we perform a time series decomposition into three components: trend, seasonal, and residual.
Our analysis focuses on the trend component, which we model using five specifications within a linear mixed-effects modeling framework, all yielding consistent results.
All five models include time as a fixed effect and estimate an intercept and a slope associated with this variable.
The intercept is interpreted as the magnitude of the initial drop in mobility, while the slope represents the rate at which mobility levels recover over time.
Models 3, 4, and 5 incorporate random effects to capture variation in the intercept, slope, or both, based on the category of the origin of the flows.
Detailed descriptions of all model specifications are provided in Methodology Section X.

Figures @fig-redensity and @fig-rerdi present, for each country, the random variations in intercept and slope estimated by Models 3, 4, and 5, providing insights into the drivers of intergroup variation in the recovery towards baseline mobility levels.
Focusing first on @fig-redensity, we observe that

![Figure random effects density](figures/Fig-re-density.pdf){#fig-redensity}

![Figure random effects RDI](figures/Fig-re-rdi.pdf){#fig-rerdi}

## Focus on cities {#sec-cities}

![FUAS netflows](figures/Fig-fuas-netflows.pdf){#fig-cities}
